\vspace{2cm} \LARGE \bf Seasonal influenza circulation patterns and projections for 2017-2018

### 2017-09-19 auspice/reports/tex/sep-2017.tex
Textual changes

\textbf{H3N2 continues to diversify with many coexisting clades, all of
which carry several amino acid mutations at previously characterized
epitopes. The majority of viruses fall into the 3c2.a clade which has
been dominating globally for \textgreater{}3 years, but 3c3.a viruses
continue to persist. The common ancestor of circulating H3N2 viruses is
now more than 5 years old, which is rare for H3N2. Despite extensive
genetic diversity, serological assays suggest limited, but non-zero,
antigenic evolution. We expect multiple competing clades within 3c2.a to
persist into the future with no clear immediate winner.}

@@

We observe substantial diversification within the 3c2.a clade over the
past year. There are 5 major subclades within 3c2.a that have reached
appreciable frequency (\FIG{h3n2_tree_with_timetree}). These clades are characterized by the following
mutations and highlighed in the phylogenetic tree in \FIG{h3n2_tree}:

@@

All of these clades possess mutations at previously characterized
epitope sites.

All of these clades have risen in frequency and are now competing with
one another (\FIG{h3n2_mutations}, \FIG{h3n2_mutations_by_region}). Clade 1 (53N, 144R, 171K, 192T, 197H) is highly derived in
terms of sequence, but is at low frequency globally (currently less than
1\%) despite making up the a substantial fraction of 2017 Oceania
viruses. Clade 2 (121K, 144K) had a large upswing around Dec 2016, but has since declined slightly to $\sim$21\% global frequency. Clade 3 (131K, 142K,
261Q) and clade 4 (135K, HA2:150E) have each slowly climbed to
approximately 20\% frequency. However, clade 5 (92R, 311Q) has recently
increased to 39\% in global frequency.

@@

All 5 of these clades appear somewhat drifted according to antigenic
analysis \cite{neher2015prediction} of HI measurements provided by the Influenza Division at the US CDC (\FIG{h3n2_tree_titer_model}).
Antigenic differences from vaccine strain A/HongKong/4801/2014 are estimated to be between 0.5 and 2.0 antigenic units (two-fold HI dilutions).

The mutation T135K has appeared independently in clades 2, 4 and 5 and is now globally at 40\% frequency.
The most rapidly growing clade carries mutations
N121K/K92R/H311Q/E62G/R142G/T135K. This is a subclade of clade 5 above
with addition of 62G, 142G and 135K. The model fit to the antigenic data
estimates that this subclade of clade 5 has drifted by 2.2 antigenic
units, 1 of which is derived from the E62G/R142G mutations and 1.2 of which is derived from
the T135K mutation (on this background), see \FIG{h3n2_tree_135k}.
Note, however, that there
is no evidence of antigenic evolution in the FRA data (one data point
only). A small sister clade with mutation T135N (also part of clade 5)
recently developed. Taken together, changes at position 135 likely have
selective origin.
Previously, the mutation K135T was associated with the Sidney/97 antigenic transition as an ``accessory'' mutation \cite{koel2013substitutions}.

@@

Clade 3 with T131K/R142K mutations continues to be common in China (but
not in South Asia) with an embedded cluster of South and North American
isolates from last summer. Subclades within this clade have been
expanding recently, but there is little evidence for antigenic evolution
in the HI or FRA data. The 131K substitution is not seen outside this
clade.

@@

Additionally, an analysis of ``local branching index'' \cite{neher2014predicting} suggests that the E62G/R142G/T135K subclade within clade 5 is the most rapidly growing of circulating viruses and is thus estimated to have the highest fitness.
With the exception of background 3c2.a1 viruses, other circulating 3c2.a are seen to have similar local branching index values.

@@

\textit{In summary, we think that the clade most likely to increase further in frequency is the E62G/R142G/T135K subclade within clade 5.
However, clade 2 viruses (characterized by N121K/R144K) and clade 3 viruses (characterized by T131K/R142K/R261Q) also appear competitive (they are present at moderate frequency, have somewhat diverged antigenic phenotypes and are increasing in frequency), although less so than E62G/R142G/T135K viruses (which are growing faster and appear more diverged in antigenic phenotype).
We expect all three of these clades to contribute to the future H3N2 population.}

>>>>

\textbf{H3N2 continues to diversify with many coexisting clades, all of
which carry several amino acid mutations at previously characterized
epitope sites. The majority of viruses fall into the 3c2.a clade which has
been dominating globally for \textgreater{}3 years, but 3c3.a viruses
continue to persist. The common ancestor of circulating H3N2 viruses is
now more than 5 years old, which is rare for H3N2. Despite extensive
genetic diversity, serological assays suggest limited, but non-zero,
antigenic evolution. We expect multiple competing clades within 3c2.a to
persist into the future with no clear immediate winner.}

@@

We observe substantial diversification within the 3c2.a clade over the
past year. There are 5 major subclades within 3c2.a that have reached
appreciable frequency (\FIG{h3n2_tree_with_timetree}). All of these clades possess mutations at previously characterized
epitope sites. These clades are characterized by the following
mutations and highlighed in the phylogenetic tree in \FIG{h3n2_tree_with_timetree}:

@@

All of these clades have risen in frequency and are now competing with
one another (\FIG{h3n2_mutations}, \FIG{h3n2_mutations_by_region}). Clade 1 (53N, 144R, 171K, 192T, 197H) is highly derived in
terms of sequence, but is at low frequency globally (currently less than
1\%) despite making up the a substantial fraction of 2017 Oceania
viruses.
Clade 2 (121K, 144K) had a large upswing around Dec 2016, but has since declined slightly to $\sim$27\% global frequency, but is now increasing again.
Clade 3 (131K, 142K, 261Q) has steadily climbed to $\sim$39\% global frequency and importantly is at relatively high frequency in Asia.
Clade 4 (135K, HA2:150E) has persisted at low frequency and is currently at $\sim$8\% globally.
Clade 5 (92R, 311Q) has recently increased to $\sim$25\% in global frequency, also with relatively high frequency in Asia.

@@

All 5 of these clades appear somewhat drifted according to antigenic
analysis \cite{neher2015prediction} of HI measurements provided by the Influenza Division at the US CDC (\FIG{h3n2_tree_titer_model}).
Antigenic differences from vaccine strain A/HongKong/4801/2014 are estimated to be between 0.5 and 2.2 antigenic units (or up to a four-fold HI dilution).

The mutation T135K has appeared independently in clades 2, 4 and 5 and is now globally at 31\% frequency.
The most rapidly growing clade carries mutations
N121K/K92R/H311Q/E62G/R142G/T135K. This is a subclade of clade 5 above
with addition of 62G, 142G and 135K. The model fit to the antigenic data
estimates that this subclade of clade 5 has drifted by 2.2 antigenic
units, 1 of which is derived from the E62G/R142G mutations and 1.2 of which is derived from
the T135K mutation (on this background), see \FIG{h3n2_tree_135k}.
Note, however, that there
is no evidence of antigenic evolution in the FRA data (one data point
only). A small sister clade with mutation T135N (also part of clade 5)
recently developed. Taken together, changes at position 135 likely have
selective origin.
Previously, the mutation K135T was associated with the Sidney/97 antigenic transition as an ``accessory'' mutation \cite{koel2013substitutions}.

@@

Clade 3 with T131K/R142K mutations continues to be common in China (but
not in South Asia) with an embedded cluster of South and North American
isolates from last summer. Subclades within this clade have been
expanding recently, and there is evidence for a small amount of antigenic evolution
in the HI (0.9 antigenic units) and FRA data (0.1 antigenic units). The 131K substitution is not seen outside this
clade.

@@

Additionally, an analysis of ``local branching index'' \cite{neher2014predicting} suggests that the E62G/R142G/T135K subclade within clade 5 is the most rapidly growing of circulating viruses and is thus estimated to have the highest fitness.
With the exception of background 3c2.a1 viruses, other circulating 3c2.a viruses are seen to have similar local branching index values.

@@

\textit{In summary, we think that the clade most likely to increase further in frequency is the E62G/R142G/T135K subclade within clade 5.
However, clade 2 viruses (characterized by N121K/R144K) and clade 3 viruses (characterized by T131K/R142K/R261Q) also appear competitive (they are present at moderate frequency, have slightly diverged antigenic phenotypes and are increasing in frequency), although less so than E62G/R142G/T135K viruses (which are growing faster and appear more diverged in antigenic phenotype).
We expect all three of these clades to contribute to the future H3N2 population.}
